---
title: Generalised Linear Models 2
author: Bongani Ncube
date: '2023-09-09'
slug: generalised-linear-models-2
categories:
  - classification
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2023-09-09T12:49:23+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
<script src="//yihui.org/js/math-code.js" defer></script>
<!-- Just one possible MathJax CDN below. You may use others. -->
<script defer
  src="//mathjax.rstudio.com/latest/MathJax.js?config=TeX-MML-AM_CHTML">
</script>

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, 
                      echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.path = "static",
                      fig.height=6, 
                      fig.width = 1.777777*6,
                      fig.align='center',
                      tidy = FALSE, 
                      comment = NA, 
                      highlight = TRUE, 
                      prompt = FALSE, 
                      crop = TRUE,
                      comment = "#>",
                      collapse = TRUE)
knitr::opts_knit$set(width = 60)
library(tidyverse)
library(reshape2)
theme_set(theme_light(base_size = 16))
make_latex_decorator <- function(output, otherwise) {
  function() {
      if (knitr:::is_latex_output()) output else otherwise
  }
}
insert_pause <- make_latex_decorator(". . .", "\n")
insert_slide_break <- make_latex_decorator("----", "\n")
insert_inc_bullet <- make_latex_decorator("> *", "*")
insert_html_math <- make_latex_decorator("", "$$")
```

# Multinomial

If we have more than two categories or groups that we want to model relative to covariates (e.g., we have observations $i = 1,\cdots,n$ and groups covariates $j = 1,\cdots,J$), multinomial is our candidate model

Let

-   $p_{ij}$ be the probability that the i-th observation belongs to the j-th group
-   $Y_{ij}$ be the number of observations for individual i in group j; An individual will have observations $Y_{i1},Y_{i2},…Y_{iJ}$
-   assume the probability of observing this response is given by a multinomial distribution in terms of probabilities $p_{ij}$, where $\sum_{j = 1}^J p_{ij} = 1$ . For interpretation, we have a baseline category $p_{i1} = 1 - \sum_{j = 2}^J p_{ij}$

#
The link between the mean response (probability) $p_{ij}$ and a linear function of the covariates

$$
\eta_{ij} = \mathbf{x'_i \beta_j} 
$$

which equals

$$\log \frac{p_{ij}}{p_{i1}}, j = 2,..,J
$$

We compare $p_{ij}$ to the baseline $p_{i1}$, suggesting

$$
p_{ij} = \frac{\exp(\eta_{ij})}{1 + \sum_{i=2}^J \exp(\eta_{ij})}
$$

which is known as **multinomial logistic** model.

# Note:

-   Softmax coding for multinomial logistic regression: rather than selecting a baseline class, we treat all K class symmetrically - equally important (no baseline).

$$
P(Y = k | X = x) = \frac{exp(\beta_{k1} + \dots + \beta_{k_p x_p})}{\sum_{l = 1}^K exp(\beta_{l0} + \dots + \beta_{l_p x_p})}
$$
then the log odds ratio between $k-th$ and $k^{t}th$ classes is

$$
\log (\frac{P(Y=k|X=x)}{P(Y = k' | X=x)}) = (\beta_{k0} - \beta_{k'0}) + \dots + (\beta_{kp} - \beta_{k'p}) x_p
$$

# Poisson GLMs

### The Poisson distribution

The Poisson distribution specifies the probability of a
discrete random variable $Y$ and is given by:


$$f(y, \,\mu)\, =\, Pr(Y = y)\, =\, \frac{\mu^y \times e^{-\mu}}{y!}$$

$$E(Y)\, =\, Var(Y)\, =\, \mu$$

#
where $\mu$ is the parameter of the Poisson distribution

The Poisson distribution is particularly relevant to model count data because it:

-   specifies the probability only for integer values
-   $P(y<0) = 0$, hence the probability of any negative value is null
-   $Var(Y) = \mu$ (the mean-variance relationship) allows for heterogeneity (e.g. when
    variance generally increases with the mean)

A Poisson GLM will model the value of $\mu$ as a function of different explanatory variables:

# **Step 1.**

We assume $Y_i$ follows a Poisson distribution with mean and variance $\mu_i$.

$$Y_i \sim Poisson(\mu_i)$$

$$E(Y_i) = Var(Y_i) = \mu_i$$

$$f(y_i, \, \mu_i) = \frac{\mu^{y_i}_i \times e^{-\mu_i}}{y!}$$

$\mu_i$ corresponds to the expected number of individuals.

# **Step 2.**

We specify the linear predictor of the model just as in a linear model.

$$\underbrace{\alpha}_\text{Intercept} + \underbrace{\beta_1}_\text{slope of 'variable'} \times \text{variable}_i$$

# **Step 3.**

The link between the mean of $Y_i$ and linear predictor is a logarithmic function can be written as:

$$log(\mu_i) = \alpha + \beta_1 \times \text{variable}_i $$

It can also be written as:

$$\mu_i = e^{ \alpha + \beta \times \text{variable}_i}$$

This shows that the impact of each explanatory variable is multiplicative.
Increasing variable by one increases $\mu$ by factor of exp( $\beta_\text{variable}$ ).

We can also write it as:

$$\mu_i = e^{\alpha} \times e^{\beta_1^{\text{variable}_i}}$$

If $\beta_j = 0$ then $exp(\beta_j) = 1$ and $\mu$ is not related to $x_j$. If $\beta_j > 0$ then $\mu$ increases if $x_j$ increases; if $\beta_j < 0$ then $\mu$ decreases if $x_j$ increases.

# setup
```{r}
df<-readxl::read_xlsx("multinomial.xlsx")
df<-df |> 
  separate(`y,party,race,gender`,into=c("y","party","race","gender"),",") |> 
  mutate_if(is.character,as.factor) |> 
  mutate(y=as.numeric(y)) |> 
  rename(income="y")
df

write_csv(df,"multinom.csv")
```

#
```{r}
summary(df)
```
#
```{r}
table(df$party,df$race)
```

# model with all variables

```{r}
model<-nnet::multinom(party~.,data=df)
model
```

#
```{r}
summary(model)
```

#

$$\log(\frac{\pi_I}{\pi_D})=12.33053-1.55income-6.07377raceWhite-0.07011659genderMale$$
$$\log(\frac{\pi_R}{\pi_D})=1.12294-0.113income-0.52094raceWhite-0.124genderMale$$
```{r}
p1<-exp(2.181+11.67-0.096*5000)/(1+exp(2.181+11.67-0.096*5000)+exp(1.58+6.55-0.053*5000))
p1
```

```{r}
p2<-exp(1.58+6.55-0.053*5000)/(1+exp(2.181+11.67-0.096*5000)+exp(2.181+11.67-0.096*5000))
p2
```

```{r}
1-p1-p2
```

#
```{r}
odds<-exp(1.12294-0.1137472*400)
odds
```


# NULL MODEL
```{r}
mod_null<-nnet::multinom(party~1,data=df)
mod_null
```

# model with gender
```{r}
(model_gend<-nnet::multinom(party~gender,data=df))

```

# compare the models
```{r}
anova(mod_null,model_gend,test="Chisq")
```

# model with race
```{r}
(model_race<-nnet::multinom(party~race,data=df))
```

# cmpare race model to null model
```{r}
anova(mod_null,model_race,test="Chisq")
```

# model with income
```{r}
(model_inc<-nnet::multinom(party~income,data=df))
```

# compare income model to null model
```{r}
anova(mod_null,model_inc,test="Chisq")
```

# model with race and income
```{r}
(model_inc_race<-nnet::multinom(party~income+race,data=df))
```

# compare models
```{r}
anova(model_inc_race,model_inc,test="Chisq")
```

# check using chi square distribution
```{r}
chisq.test(table(df$party,df$race))
```

# poisson dataset

```{r}
df<-read.table("canada.txt",sep="",header=TRUE) |> 
  mutate_if(is.character,as.factor) |> 
  as_tibble()
df
```

#
```{r}
ggplot(df,aes(x=dead))+
  geom_density()
```

# model with all variables
```{r}
model<-glm(dead~age+smoke+pop,data=df,family = poisson)
model
```

#
```{r,eval=FALSE}
extract_eq(model,terms_per_line = 2,wrap=TRUE,use_coefs = TRUE, greek_colors = "blue")
```
$$
\begin{aligned}
\log ({ \widehat{E( \operatorname{dead} )} })  &= 2.77 + 0.62(\operatorname{age}_{\operatorname{45-59}})\ + \\
&\quad 1(\operatorname{age}_{\operatorname{50-54}}) + 1.38(\operatorname{age}_{\operatorname{55-59}})\ + \\
&\quad 1.5(\operatorname{age}_{\operatorname{60-64}}) + 2.14(\operatorname{age}_{\operatorname{65-69}})\ + \\
&\quad 2.36(\operatorname{age}_{\operatorname{70-74}}) + 2.2(\operatorname{age}_{\operatorname{75-79}})\ + \\
&\quad 1.95(\operatorname{age}_{\operatorname{80+}}) + 0.57(\operatorname{smoke}_{\operatorname{cigarretteOnly}})\ + \\
&\quad 0.29(\operatorname{smoke}_{\operatorname{cigarrettePlus}}) - 0.21(\operatorname{smoke}_{\operatorname{no}})\ + \\
&\quad 0(\operatorname{pop})
\end{aligned}
$$

# model_null
```{r}
mod1<-glm(dead~1,data=df,family = poisson)
```
# model with age
```{r}
mod2<-glm(dead~age,data=df,family = poisson)
```

# COMPARE THE MODELS
```{r}
anova(mod1,mod2,test="LRT")
```

# model with population
```{r}
mod3<-glm(dead~pop,data=df,family = poisson)
```

# COMPARE NULL TO POPULATION MODEL
```{r}
anova(mod1,mod3,test="LRT")
```

# model with smoking
```{r}
mod4<-glm(dead~smoke,data=df,family = poisson)
```

# COMPARE SMOKING MODEL TO NULL MODEL
```{r}
anova(mod1,mod4,test="LRT")
```

# REPORT BEST MODEL
```{r}
report::report(model)
```


