---
title: Control flows in R
author: Bongani Ncube
date: '2023-09-09'
slug: control-flows-in-r
categories:
  - Programming
  - Rstats
  - introduction
  - classification
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2023-09-09T12:16:40+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
<script src="//yihui.org/js/math-code.js" defer></script>
<!-- Just one possible MathJax CDN below. You may use others. -->
<script defer
  src="//mathjax.rstudio.com/latest/MathJax.js?config=TeX-MML-AM_CHTML">
</script>

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, 
                      echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.path = "static",
                      fig.height=6, 
                      fig.width = 1.777777*6,
                      fig.align='center',
                      tidy = FALSE, 
                      comment = NA, 
                      highlight = TRUE, 
                      prompt = FALSE, 
                      crop = TRUE,
                      comment = "#>",
                      collapse = TRUE)
knitr::opts_knit$set(width = 60)
library(tidyverse)
library(reshape2)
theme_set(theme_light(base_size = 16))
make_latex_decorator <- function(output, otherwise) {
  function() {
      if (knitr:::is_latex_output()) output else otherwise
  }
}
insert_pause <- make_latex_decorator(". . .", "\n")
insert_slide_break <- make_latex_decorator("----", "\n")
insert_inc_bullet <- make_latex_decorator("> *", "*")
insert_html_math <- make_latex_decorator("", "$$")
```

Every time some operations have to be repeated, a *loop* may come in handy. Loops are good for:

- Doing something for every element of an object;
- Doing something until the processed data runs out;
- Doing something for every file in a folder;
- Doing something that can fail, until it succeeds;
- Iterating a calculation until it reaches convergence.

# in this series we look at :

+ `for loops`
+ We will also dive into the `apply()` family functions, which are interesting function-based    alternatives to `for() {}` loops.

# lets load some data for later on;

```{r}
# take a few rows
loan_data<-readr::read_csv("loan_data_cleaned.csv") |> 
  select(-1) |> 
  sample_n(size=15) |> 
  as.data.frame()

```


# `for` loops

A `for()` loop works in the following way:

```r
for(i in sequence) {
expression
}
```

#
```{r flowchart5for-loop, echo = FALSE, purl=FALSE}
library(DiagrammeR)

#Create node data frame
f5_nodes <- create_node_df(n = 4, #number of nodes
                           type = 'a', #for grouping
                           label = c(" ", "Last item\nreached?", "expression\nfor() \nbody", "  "), 
                           shape = c("circle", "diamond", 'rectangle', 'circle'), 
                           style = c('solid', 'solid', 'solid', 'filled'), 
                           fontname = 'Helvetica',
                           fontsize = 10,
                           fixedsize = FALSE,
                           color = 'mediumblue',
                           width = .1
)


#create edges
f5_edges <- create_edge_df(from = c(1, 2, 2, 3), #origin node id
                           to = c(2, 3, 4, 2), 
                           label = c(' ', ' if FALSE', '  if TRUE, exit loop', ' '), 
                           fontsize = 10,
                           color = 'dimgrey',
                           decorate = TRUE,
                           tailport = c('s', 'w', 'e', 'e'),
                           headport = c('n', 'w', 'n', 's')
)

#create flowchart
flowchart5 <- create_graph(
  nodes_df = f5_nodes,
  edges_df = f5_edges,
  directed = TRUE, 
  attr_theme = 'tb' 
)

#display flowchart
render_graph(flowchart5, width = '35%', height = 'auto')
```

# note

The letter `i` can be replaced with any variable name, `sequence` can be elements or the position of these elements, and `expression` can be anything. Try the examples below:

# example 1

```{r, echo = TRUE}
for(a in c("Hello", 
           "R/Posit", 
           "group members")) {
  print(a)
}
```

# example 2
```{r, echo = TRUE}
for(z in 1:4) {
  a <- rnorm(n = 1, 
             mean = 5 * z, 
             sd = 2)
  print(a)
}
```

# example 3

In this next example, every instance of `m` is being replaced by each number between `1` and `7`, until it reaches the last element of the sequence

```{r eval = TRUE, echo = TRUE}
y <- 2
for(m in 1:6) {
  print(y*m)
}
```


# `for` loops on different classes

As expected, you can use `for()` loops in different object types and classes, such as a `list`. Let us take the example below, where we are creating the `elements` object list.

```{r echo = TRUE}
(elements <- list(a = 1:3, 
                  b = 4:10, 
                  c = 7:-1))
```
#
Now, let us `print` the double of every element of the list:

```{r echo = TRUE}
for(element in elements) {
  print(element*2)
}

```

# `for` and `if` together

```{r, echo = TRUE, eval = TRUE}
x <- c(2, 5, 3, 9, 6,8)
count <- 0
```



```{r, echo = TRUE, eval = TRUE}
for(val in x) {
  if(val %% 2 == 0) {
    count <- count + 1
  }
}
print(count)

## basically this counts the number of even numbers(numbers divisible by 2)
```


# `for` with a real dataset

`for()` loops are often used to loop over a dataset. We will use loops to perform functions on the `loan_data` . To load and see the first 6 rows of the `loan data` dataset, execute the following code:

```{r, echo = TRUE, eval = TRUE}
head(loan_data)
```
# Now, to recursively print loan amount, let us do this: 

```{r, echo = TRUE}
for(i in 1:length(loan_data[,1])) { # for each row in the loan_data dataset
  print(loan_data$loan_amnt[i]) # print the loan amount
}
```


# first five rows

```{r,purl=FALSE}
for(i in 1:5) { 
  print(loan_data$loan_amnt[i]) 
}
```
# what about last 5
```{r, purl=FALSE}
for (i in 11:15) { 
  print(loan_data$loan_amnt[i]) 
}
```

# Now, let us obtain the loan amount for defaulters only 

```{r, purl=FALSE}
for(i in 1:length(loan_data[,1])) { # for each row in the loan_data dataset
  if(loan_data$loan_status[i] == 1) { # if the type is "0"
    print(loan_data$loan_amnt[i]) # print the loan_amount
  }
}
```

# To loop over the number of rows of a data frame, we can use the function `nrow()`:

```{r, echo = TRUE}
for(i in 1:nrow(loan_data)) {
  # for each row in
  # the loan_data dataset
  print(loan_data$loan_amnt[i])
  # print the loan_amount
}
```

# To perform operations on the elements of one column, we can directly iterate over it.

```{r, echo = TRUE}
for(p in loan_data$loan_amnt) {
  # for each element of
  # the column "conc" of
  # the loan_data df
  print(p)
  # print the p-th element
}
```

#
The expression within the loop can be almost anything and is usually a compound statement containing many commands.

```{r, echo = TRUE}
for(i in c(2,5:6)) { # for i in 4 to 5
  print(colnames(loan_data)[i])
  print(mean(loan_data[,i])) # print the mean of that column from the loan_data dataset
}
```


# The `apply()` family

`R` disposes of the `apply()` function family, which consists of iterative functions that aim at **minimizing your need to explicitly create loops**.

## `apply()`

Let us consider that we have a `height` matrix containing the height (in metres) that was taken from five individuals (in rows) at four different times (as columns). 

```{r, echo = TRUE, eval = TRUE}
(height <- matrix(runif(20, 1.5, 2),
                  nrow = 5,
                  ncol = 4))
```

## We would like to obtain the average height at each time step.

One option is to use a `for() {}` loop to iterate from column `1` to `4`, use the function `mean()` to calculate the average of the values, and sequentially store the output value in a vector.

Alternatively, we can use the `apply()` function to set it to apply the `mean()` function to every column of the `height` matrix. See the example below:

#
```{r, echo = TRUE, eval = TRUE}
apply(X = height,
      MARGIN = 2,
      FUN = mean)
```

> The `apply()` function begins with three arguments main arguments: `X`, which will take a matrix or a data frame; `FUN`, which can be any function that will be applied to the `MARGIN`s of `X`; and `MARGIN` which will take `1` for row-wise computations, or `2` for column-wise computations. 

# `lapply()`

`lapply()` applies a function to every element of a `list`.

The output returned is also `list` (explaining the “`l`” in `lapply`) and has the same number of elements as the object passed to it.

```{r, echo = TRUE, eval = TRUE}
SimulatedData <- list(
  SimpleSequence = 1:4,
  Norm10 = rnorm(10),
  Norm20 = rnorm(20, 1),
  Norm100 = rnorm(100, 5)
)
```

#
```{r, echo = TRUE, eval = TRUE}
# Apply mean to each element of the list

lapply(X = SimulatedData, 
       FUN = mean)
```

> `lapply()` operations done in objects different from a `list` will be coerced to a `list` via `base::as.list()`.

# `sapply()`

`sapply()` is a 'wrapper' function for `lapply()`, but returns a simplified output as a `vector`, instead of a `list`.

```{r, echo = TRUE, eval = TRUE}
SimulatedData <- list(SimpleSequence = 1:4,
                      Norm10 = rnorm(10),
                      Norm20 = rnorm(20, 1),
                      Norm100 = rnorm(100, 5))

# Apply mean to each element of the list
sapply(SimulatedData, mean)
```

# `mapply()`

`mapply()` works as a multivariate version of `sapply()`.

It will apply a given function to the first element of each argument first, followed by the second element, and so on. For example:

```{r, echo = TRUE, eval = TRUE}
lilySeeds <- c(80, 65, 89, 23, 21)
poppySeeds <- c(20, 35, 11, 77, 79)

# Output
mapply(sum, lilySeeds, poppySeeds)
```

# `tapply()`

`tapply()` is used to apply a function over subsets of a vector.

It is primarily used when the dataset contains dataset contains different groups (*i.e.* levels/factors) and we want to apply a function to each of these groups.

```{r, echo = TRUE, eval = TRUE}
# get the mean loan_amnt by grade
tapply(loan_data$loan_amnt, loan_data$grade, FUN = mean)
```
