---
title: "Visualizing data layer by layer in R"
author: 'Bongani Ncube : statistical analyst'
date: "July 2023"
output:
  beamer_presentation:
    theme: Copenhagen
    colortheme: beaver
    fonttheme: structurebold
    fig_caption: no
    includes:
      in_header: header.tex
    latex_engine: pdflatex
    slide_level: null
  pdf_document: default
  revealjs::revealjs_presentation:
  html_document:
    df_print: paged
  ioslides_presentation: default
  slidy_presentation: default
subtitle: NCUBE ON DATA
classoption: aspectratio=169
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, 
                      echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.height=6, 
                      fig.width = 1.777777*6,
                      tidy = FALSE, 
                      comment = NA, 
                      highlight = TRUE, 
                      prompt = FALSE, 
                      crop = TRUE,
                      comment = "#>",
                      collapse = TRUE)
knitr::opts_knit$set(width = 60)
library(tidyverse)
library(reshape2)
theme_set(theme_light(base_size = 16))
make_latex_decorator <- function(output, otherwise) {
  function() {
      if (knitr:::is_latex_output()) output else otherwise
  }
}
insert_pause <- make_latex_decorator(". . .", "\n")
insert_slide_break <- make_latex_decorator("----", "\n")
insert_inc_bullet <- make_latex_decorator("> *", "*")
insert_html_math <- make_latex_decorator("", "$$")
```

# background

> Visualisation is the process of representing data graphically and interacting with these representations. The objective is to gain insight into the data.

To work with **ggplot2**\index{ggplot2}, remember that at least your R codes must

-   start with `ggplot()`
-   identify which data to plot `data = Your Data`
-   state variables to plot for example `aes(x = Variable on x-axis, y = Variable on y-axis )` for bivariate
-   choose type of graph, for example `geom_histogram()` for histogram, and `geom_points()` for scatterplots\index{Scatterplot}

# setup

+ installing *tidyverse* package which contains *dplyr* and *ggplot2*

```{r,fig.height=3}
library(tidyverse)
loan_data <- read_csv('loan_data_cleaned.csv')
loan_data <- loan_data |>
  mutate(default=ifelse(loan_status==1,"defaulted","not defaulted")) |> 
  mutate_if(is.character,as.factor)
```

# getting started

+ Calling ggplot() along just creates a blank plot
```{r,fig.height=3}
ggplot()
```

# next up

+ I need to tell ggplot what data to use
```{r,fig.height=3}
ggplot(data=loan_data)
```

# grammar of graphics

+ And then give it some instructions using the grammar of graphics.
+ Let's build a simple scatterplot with annual income on the x-axis and loan amount on the y axis

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_point(mapping=aes(x=annual_inc, y=loan_amnt))
```

# refining...

+ Let's try representing a different dimension.  
+ What if we want to differentiate public vs. private schools?
+ We can do this using the shape attribute

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_point(mapping=aes(x=annual_inc, y=loan_amnt, shape=default))
```

# not neat!!..try color.

+ That's hard to see the difference.  What if we try color instead?
```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_point(mapping=aes(x=annual_inc, y=loan_amnt, color=default))
```

# try size!

+ I can also alter point size.  Let's do that to represent grade
```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_point(mapping=aes(x=annual_inc, y=loan_amnt, color=default, size=grade))
```

# transparency

+ And, lastly, let's add some transparency so we can see through those points a bit
+ Experiment with the alpha value a bit.

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_point(mapping=aes(x=annual_inc, y=loan_amnt, color=default, size=grade), alpha=1)
```

# very transparent

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_point(mapping=aes(x=annual_inc, y=loan_amnt, color=default, size=grade), alpha=1/100)
```

# How many people defaulted?

```{r,fig.height=3}
# This calls for a bar graph!
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=default))
```

# Break it out by grade

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=default, color=grade))
```

# Well, that's unsatisfying!  Try fill instead of color

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=default, fill=grade))
```

# How about loan amount by default status?

+ First, I'll use some dplyr to create the right tibble
```{r,fig.height=3}
loan_data %>%
  group_by(default) %>%
  summarize(average_amount=mean(loan_amnt))
```

# And I can pipe that straight into ggplot

+ but this will produce an error

```{r,eval=FALSE,fig.height=4}
loan_data %>%
  group_by(default) %>%
  summarize(average_amount=mean(loan_amnt)) %>%
  ggplot() +
  geom_bar(mapping=aes(x=default, y=average_amount))
```

# use geom_col() instead

+ But I need to use a column graph instead of a bar graph to specify my own y

```{r,fig.height=3}
loan_data %>%
  group_by(default) %>%
  summarize(average_amount=mean(loan_amnt)) %>%
  ggplot() +
  geom_col(mapping=aes(x=default, y=average_amount))
```

# Histograms

+ Let's look at annual income
+ Histograms can help us by binning results

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_histogram(mapping=aes(x=annual_inc), origin=0)
```

# What if we want fewer groups? Let's ask for 4 bins

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_histogram(mapping=aes(x=annual_inc), bins=4, origin=0)
```

# Or 10 bins.

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_histogram(mapping=aes(x=annual_inc), bins=10, origin=0)
```

# Or we can specify the width of the bins instead

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_histogram(mapping=aes(x=annual_inc), binwidth=1000, origin=0)
```

# large binwidth

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_histogram(mapping=aes(x=annual_inc), binwidth=10000, origin=0)
```

# changing background

## Change the plot background color
```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=loan_status, fill=home_ownership)) +
  theme(plot.background=element_rect(fill='purple'))
```

# Change the panel background color

```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=loan_status, fill=home_ownership)) +
  theme(panel.background=element_rect(fill='purple'))
```

# Let's be minimalist and make both backgrounds disappear
```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=loan_status, fill=home_ownership)) +
  theme(panel.background=element_blank()) +
  theme(plot.background=element_blank())
```

# Add grey gridlines
```{r,fig.height=3}
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=loan_status, fill=home_ownership)) +
  theme(panel.background=element_blank()) +
  theme(plot.background=element_blank()) +
  theme(panel.grid.major=element_line(color="grey"))
```

# Only show the y-axis gridlines
```{r,fig.height=3,fig.width=9}
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=loan_status, fill=home_ownership)) +
  theme(panel.background=element_blank()) +
  theme(plot.background=element_blank()) +
  theme(panel.grid.major.y=element_line(color="grey"))+
  scale_fill_manual(values=c("orange","blue","green","blue4"))
```
```{r}
library(ggthemes)
ggplot(data=loan_data) +
  geom_bar(mapping=aes(x=loan_status, fill=home_ownership)) +
  theme(panel.background=element_blank()) +
  theme(plot.background=element_blank()) +
  theme(panel.grid.major.y=element_line(color="grey"))+
  scale_fill_manual(values=c("orange","blue","green","blue4"))+
  theme_solarized()+
  theme_wsj()



```

```{r}
ggplot(data=loan_data, aes(x = loan_status)) + 
  geom_bar(fill = "chartreuse") + 
  theme(axis.text.x = element_text(angle = 90))

```

