---
title: Working with Factor Variables in R
author: Bongani Ncube
date: '2023-09-09'
slug: working-with-factor-variables-in-r
categories: []
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: '2023-09-09T13:11:50+02:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
<script src="//yihui.org/js/math-code.js" defer></script>
<!-- Just one possible MathJax CDN below. You may use others. -->
<script defer
  src="//mathjax.rstudio.com/latest/MathJax.js?config=TeX-MML-AM_CHTML">
</script>

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, 
                      echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.path = "static",
                      fig.height=6, 
                      fig.width = 1.777777*6,
                      fig.align='center',
                      tidy = FALSE, 
                      comment = NA, 
                      highlight = TRUE, 
                      prompt = FALSE, 
                      crop = TRUE,
                      comment = "#>",
                      collapse = TRUE)
knitr::opts_knit$set(width = 60)
library(tidyverse)
library(reshape2)
theme_set(theme_light(base_size = 16))
make_latex_decorator <- function(output, otherwise) {
  function() {
      if (knitr:::is_latex_output()) output else otherwise
  }
}
insert_pause <- make_latex_decorator(". . .", "\n")
insert_slide_break <- make_latex_decorator("----", "\n")
insert_inc_bullet <- make_latex_decorator("> *", "*")
insert_html_math <- make_latex_decorator("", "$$")
```


# Factors - what are they

- factors are used to work with categorical variables, variables
that  have  a  fixed  and  known  set  of  possible  values.  They  are  also
useful  when  you  want  to  display  character  vectors  in  a  non alphabetical order.
- we encounter them when modeling (linear,GLM,GAMS,Machine learning)
- handling them is on3 stage of machine learning feature engineering

# `factor()`

- Provides a way to provide different categories/enumerations of things without specifying a 
  priority preference. 
    - IE. in `loan data` grade may be a included in a regression, but 
    order in which they are included in the model shouldn't impact the results. 
    Or grade A is not greater than grade B from a regression standpoint.
    
- Option for providing ordinality when it is important to sort categories
    - IE First, Second, Third, are categories, but there is an order to them
    
# creating factors

```{r}

vec <- factor(c("Hello","world","Hello","Rposit"))
vec
```

# Levels of factors are the important piece to them, and they can be edited

```{r}
factor(c("Hello","world","Hello","Rposit"), levels = c("world","Hello"))

levels(vec) <- c("Amazing","Viewers","Are")
vec

```
#
- Imagine that you have a variable that records month:

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
sort(x1)
```
- `sort` doesn’t sort in a useful way

- There are only twelve possible months, and there’s nothing saving you from typos:
```{r}
x2 <- c("Dec", "Apr", "Jam", "Mar")
```

# how can we fix this... create a list with levels
```{r}
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)

y1 <- factor(x1, levels = month_levels)
y1

sort(y1)
```

# working with factors in datasets

- i'm still a big fan of the tidyverse
- we will use the `forcats` package to get the job done
` first lets see what happens when you read in data

### readr package

- by default treats factors as characters
- you can override this by specifying `coltypes`

# setup
```{r}

# load in libraries
library(labelled)
library(tidyverse)

# read in data
df<-readr::read_csv("loan_data_cleaned.csv")
```
#
```{r}
# look at the few first rows
head(df,n=4)

# check the classes
sapply(df[1,],class)
```

# lets override that coz we need those to be factors
```{r}
df<-readr::read_csv("loan_data_cleaned.csv",col_types = cols(loan_status=col_factor(),
                                                             grade=col_factor(),
                                                             home_ownership=col_factor(),
                                                             emp_cat=col_factor())) |> 
  select(-1)

sapply(df[1,],class) # great
```

#
- that's a draw back, cos its cumbersome, lets use `mutate_if()` instead
- we might have as many variables to deal with and listing them one by one is time consuming
- we know in advance that they will be turned to characters 

```{r}
df<-readr::read_csv("loan_data_cleaned.csv") |> 
  select(-1)
df<- df |> 
  mutate_if(is.character,as.factor)

# check class
sapply(df[1,],class) # great
```

# labeling variables
```{r}
# put variable labels
labelled::var_label(df)<-list(loan_status="default status",
                    annual_inc="annual income",
                    home_ownership="house ownership",
                    loan_amnt="loan amount",
                    emp_cat="employment category")
```

# creating factors and modifying them

- lets make use of `cut` from base R to turn numerics to factors
- i find this function very useful in most cases

#
```{r}
out_new<-df |> 
  mutate(age_group=cut(age, c(min(df$age)-1, 30,45,55,65,75, max(df$age)+1), 
                        c("< 30", "30-44","45-54","55-64","65-74","75+"), right = FALSE))
# check this out
out_new |> 
  select(age,age_group)
```
# lets have glimpse at this
```{r}
table(out_new$age_group)
### or##

out_new |>
  count(age_group)

```
#
```{r ,fig.height=4}
ggplot(out_new, aes(age_group)) +geom_bar()
```

# modifying factor order
```{r}
dat_sum <- out_new |>
  group_by(age_group) |>
  summarize(
    annual_inc= mean(annual_inc, na.rm = TRUE),
    n = n()
  )
```
#
```{r ,fig.height=4}
ggplot(dat_sum, aes(annual_inc, age_group)) + geom_point()
```

# fct_reorder() 

- It is difficult to interpret this plot because there’s no overall pattern.
- We can improve it by reordering the levels of age_group using  fct_reorder() .  _ fct_reorder()  takes three arguments:
  + f , the factor whose levels you want to modify.
  + x , a numeric vector that you want to use to reorder the levels.
  
# 
```{r ,fig.height=4}
ggplot(dat_sum, aes(annual_inc, fct_reorder(age_group,annual_inc))) + geom_point()
```
- there is some increasing order now.. its more useful when there are many levels per factor

# `fct_infreq`

- you can use `fct_infreq` to order levels in
increasing frequency: this is the simplest type of reordering because
it doesn’t need any extra variables. 
```{r}
p1<-out_new |>
  mutate(age_group = age_group |> fct_infreq() |> fct_rev()) |>
  ggplot(aes(age_group)) +
    geom_bar()
```
#
```{r ,fig.height=4}
p1
```

# fct_recode()

+ The most general and powerful  tool  is  fct_recode() .It  allows  you  to  recode,  or  change,  the value of each level. For example, take  gss_cat$partyid

```{r}
out_new |> 
  mutate(age_group1=fct_recode(age_group,
                               "less than 30" ="< 30",
                               "greater than 75"="75+")) |> 
  count(age_group1)
```
# `fct_collapse()`
+ If  you  want  to  collapse  a  lot  of  levels,  `fct_collapse()`

```{r}
out_new |> 
  mutate(age_group2=fct_collapse(age_group,
                                 "30-54"=c("30-44","45-54"),
                                 "55-74"=c("55-64","65-74"))) |> 
  count(age_group2)
```

# `fct_lump()`
+ Sometimes you just want to lump together all the small groups to
make a plot or table simpler. That’s the job of  `fct_lump()`:
```{r}
out_new %>%
  mutate(age_group3= fct_lump(age_group)) %>%
  count(age_group3)
```
#
+ The default behavior is to progressively lump together the smallest
groups, ensuring that the aggregate is still the smallest group
+ Instead,  we  can  use  the  n   parameter  to  specify  how  many  groups
(excluding other) we want to keep:

#
```{r}
out_new %>%
  mutate(age_group3= fct_lump(age_group,n=3)) %>%
  count(age_group3)
```


