---
title: 'Cox Proportional Hazards and Logistic Regression using Tidymodels: Brain Cancer'
author: "Bongani Ncube"
date: "2023-09-11"
slug: "cox-proportional-hazards-and-logistic-regression-brain-cancer"
categories:
- classification
- kaplanmeier
- logistic
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: "2023-09-11T03:11:39+02:00"
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
<script src="//yihui.org/js/math-code.js" defer></script>
<!-- Just one possible MathJax CDN below. You may use others. -->
<script defer
  src="//mathjax.rstudio.com/latest/MathJax.js?config=TeX-MML-AM_CHTML">
</script>

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE, 
                      echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      fig.path = "static",
                      fig.height=6, 
                      fig.width = 1.777777*6,
                      fig.align='center',
                      tidy = FALSE, 
                      comment = NA, 
                      highlight = TRUE, 
                      prompt = FALSE, 
                      crop = TRUE,
                      comment = "#>",
                      collapse = TRUE)
knitr::opts_knit$set(width = 60)
library(tidyverse)
library(reshape2)
theme_set(theme_light(base_size = 16))
make_latex_decorator <- function(output, otherwise) {
  function() {
      if (knitr:::is_latex_output()) output else otherwise
  }
}
insert_pause <- make_latex_decorator(". . .", "\n")
insert_slide_break <- make_latex_decorator("----", "\n")
insert_inc_bullet <- make_latex_decorator("> *", "*")
insert_html_math <- make_latex_decorator("", "$$")
```

# Introduction
In this tutorial i explore the brain cancer dataset . This is a work in progress and will be continuously updated along the way . The main emphasis is in using the **tidymodels** approach for fitting regression models in R.

## Methodology

My research aims at patients’ brain cancer survival analysis . In the analysis I did a detailed research on the variables that determine whether patients get over a disease or surrender in a certain duration time. Thus I used a  non-parametric statistic estimator to create a Kaplan-Meier Survival model to measure the survival function from lifetime duration data. We fit a cox proportional model and Logistic regression model. 

## Brain Cancer Data
A data frame with 88 observations and 8 variables:

+ sex       : Factor with levels "Female" and "Male"
+ diagnosis : Factor with levels "Meningioma", "LG glioma", "HG glioma", and "Other".
+ Location  : factor with levels "Infratentorial" and "Supratentorial".
+ ki        : Karnofsky index.
+ gtv       : Gross tumor volume, in cubic centimeters.
+ stereo    : Stereotactic method factor with levels "SRS" and "SRT".
+ status    : Whether the patient is dead at the end of the study: 0=No, 1=yes.
+ time      : Survival time, in months.

# Setup

```{r}
library(pacman)
p_load(tidyverse, tidymodels, ISLR2, survival, censored)

glimpse(BrainCancer)
```

```{r}
skimr::skim(BrainCancer)
```





# Explanatory data Analysis(EDA)

## mortality Rate

```{r}
data<-BrainCancer |> 
  mutate(brain_status=ifelse(status==1,"yes","no"))


data|>
  dplyr::group_by(brain_status)|> 
  dplyr::summarize(count=n(),
             mean=mean(gtv, na.rm = T),
             std_dev=sd(gtv, na.rm = T),
             median=median(gtv, na.rm = T), 
             min=min(gtv, na.rm = T),
             max=max(gtv, na.rm = T))|> 
 tibble::column_to_rownames(var = "brain_status") |> 
 base::as.matrix()

#Calculate mortality rate
print(paste('Mortality rate (%) is ',round(35/(35+53)*100,digits=2)))
```
```{r,echo=FALSE}
blogdown::shortcode("spoiler",text="click here",.content="the Mortality rate is approximately `40%`")
```

```{r , warning=FALSE}
library(extrafont)
loadfonts(quiet=TRUE)

iv_rates <- data |>
  group_by(brain_status) |>
  summarize(count = n()) |> 
  mutate(prop = count/sum(count)) |>
  ungroup() 

plot<-iv_rates |>
  ggplot(aes(x=brain_status, y=prop, fill=brain_status)) + 
  geom_col(color="black",width = 0.5)+ 
  theme(legend.position="bottom") + 
  geom_label(aes(label=scales::percent(prop)), color="white") + 
  labs(
    title = "mortality ratio",
    subtitle = "Brain Cancer Analysis",
    y = "proportion(%)", 
    x = "",
    fill="status",
    caption="B.Ncube::Data enthusiast") + 
  scale_y_continuous(labels = scales::percent)+ 
geom_hline(yintercept = (sum(data$status)/nrow(data)), col = "white", lty = 2) +
  tvthemes::scale_fill_kimPossible()+
  tvthemes::theme_theLastAirbender(title.font="Slayer",
                                   text.font = "Slayer")+
  theme(legend.position = 'right')
plot
```

## mortality rate by sex
### Create a customised function for summarising  categorical data per status

```{r}
summarize_status <- function(tbl){
  tbl %>% 
  #group_by(occup) %>% 
  summarise(n_died = sum(brain_status == "yes"), n_total = n()) %>%
  ungroup() %>% 
  mutate(pct_died = n_died / n_total) %>% 
  arrange(desc(pct_died)) %>% 
  mutate(low = qbeta(0.025, n_died + 5, n_total - n_died + .5),
         high = qbeta(0.975, n_died + 5, n_total - n_died + .5),
         pct = n_total / sum(n_total),
         percentage=scales::percent(pct_died))
  } 
```

### mortality rate summary per sex

```{r}
data |> 
  group_by(sex) |> 
  summarize_status()
```

+ from the dataset , $\frac{20}{43}$ men died while the other 15 where woman who died
+ we can present these more visually with the graphs below


```{r,echo=FALSE,warning=FALSE}
iv_rates <- data |> 
  group_by(brain_status,sex) |>
  summarize(count = n()) |> 
  mutate(prop = count/sum(count)) |>
  ungroup() 

plot<-iv_rates |>
  ggplot(aes(fill=brain_status, y=prop, x=sex)) + 
  geom_col(color="black",width = 0.5,position="dodge")+ 
  theme(legend.position="bottom") + 
  geom_label(aes(label=scales::percent(prop)), color="white") + 
  labs(
    title = "Mortality ratio",
    subtitle = "Brain Cancer Analysis",
    y = "proportion(%)", 
    x = "",
    fill="status",
    caption="B.Ncube::Data enthusiast") + 
  scale_y_continuous(labels = scales::percent)+ 
geom_hline(yintercept = (sum(data$status)/nrow(data)), col = "white", lty = 2) +
  tvthemes::scale_fill_kimPossible()+
  tvthemes::theme_theLastAirbender(title.font="Slayer",
                                   text.font = "Slayer")+
  theme(legend.position = 'right')
plot
```

```{r}
library(glue)
bind_count = function(x){
  as_tibble(x) %>% 
  add_count(value) %>% 
  mutate(value = glue("{value} ({n})")) %>%
    pull(value)
  
}
```

```{r}
# Scatter plot
data %>% 
  na.omit() %>% 
  group_by(diagnosis = bind_count(diagnosis)) %>% 
  summarize_status() %>% 
  mutate(diagnosis = fct_reorder(diagnosis, pct_died)) %>% 
  ggplot(mapping = aes(x = pct_died, y = diagnosis)) +
  geom_point(aes(size = pct), show.legend = T) +
  geom_errorbarh(aes(xmin = low, xmax = high), height = .3) +
  labs(
    x = "Percentage of patients in each category who died",
    title = "Distribution of status by diagnosis",
    size = "%prevalence",
    subtitle = ""
  ) +
  scale_x_continuous(labels = percent) +
  scale_size_continuous(labels = percent) +
  theme(plot.title = element_text(hjust = 0.5))
```

{{% callout note %}}

It seems that cumulatively, people with **LG glioma** and **HG glioma**  had high chances of death.

{{% /callout %}}






## distribution of numeric variables

```{r}

# Histogram of all numeric variables
data %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~ key, scales = "free") +
  geom_histogram(bins=30,fill=tvthemes::avatar_pal()(1))+
  ggthemes::theme_wsj()
```


{{% callout note %}}

+ status is really not supposed to be numeric but rather a factor variable
+ time and gross tumor volume are all positively skewed
+ karnofsky index is as good as being a factor so i will further categorize the data 

{{% /callout %}}


## further manipulation

```{r}
data <- data |> 
  mutate(kan_index = ifelse(ki>=80,"index > 79","index < 80"))

data |> 
  janitor::tabyl(kan_index)
```

## comparing dependent variable with numeric variables

```{r}
subset <- data |>
  dplyr::select(gtv,time,ki,status)


# Bring in external file for visualisations
source('functions/visualisations.R')

# Use plot function
plot <- histoplotter(subset, status, 
                     chart_x_axis_lbl = "survival status", 
                     chart_y_axis_lbl = 'Measures',
                     boxplot_color = 'navy', 
                     boxplot_fill = tvthemes::avatar_pal()(1), 
                     box_fill_transparency = 0.2) 

# Add extras to plot
plot + 
  tvthemes::theme_theLastAirbender() +
  tvthemes::scale_color_avatar()+
  theme(legend.position = 'top') 
 

```

## categorical data

```{r}
iv_rates <- data |>
  dplyr::mutate(status = ifelse(status==1,"yes","no")) |>
  select(kan_index,loc,stereo,status) |> 
  gather() |> 
  group_by(key,value) |>
  summarize(count = n()) |> 
  mutate(prop = count/sum(count)) |>
  ungroup() 

plot<-iv_rates |>
  ggplot(aes(fill=key, y=prop, x=value)) + 
  geom_col(color="black",width = 0.5)+ 
  facet_wrap(~key,scales="free") +
  theme(legend.position="bottom") + 
  geom_label(aes(label=scales::percent(prop)), color="white") + 
  labs(
    title = "survival ratio",
    subtitle = "Brain Cancer Analysis",
    y = "proportion(%)", 
    x = "",
    fill="status",
    caption="B.Ncube::Data enthusiast") + 
  scale_y_continuous(labels = scales::percent)+
  tvthemes::scale_fill_kimPossible()+
  tvthemes::theme_theLastAirbender(title.font="Slayer",
                                   text.font = "Slayer")+
  theme(legend.position = 'none')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
plot
```


```{r}
library(gtsummary)
data |>
  select(sex,diagnosis,loc,stereo,status) |> 
  tbl_summary(
    by = status,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"), 
    label = status ~ "status")|> 
  add_p(test = all_continuous() ~ "t.test",
        pvalue_fun = function(x) style_pvalue(x, digits = 2))|> 
   modify_header(statistic ~ "**Test Statistic**")|>
  bold_labels()|> 
  modify_fmt_fun(statistic ~ style_sigfig) %>% 
  modify_caption("**Table 1. Patient Characteristic**")  %>%
  modify_header(label ~ "**Variable**") %>% 
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Survival Status**") %>%
  modify_footnote(all_stat_cols() ~ "Mean (SD) or Frequency (%)") %>%
  bold_labels() %>%
  as_gt()
```

**Comments**
+ the plimenary suggest that at 5\% level of significance ,diagnosis( Meningioma , LG glioma ,HG glioma and other diagnostics) and stereo variable are significantly associated with survival of patients.

```{r}
data |>
  select(gtv,time,sex) |> 
  tbl_summary(
    by = sex,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"), 
    label = sex ~ "sex")|> 
  add_p(test = all_continuous() ~ "t.test",
        pvalue_fun = function(x) style_pvalue(x, digits = 2))|> 
   modify_header(statistic ~ "**Test Statistic**")|>
  bold_labels()|> 
  modify_fmt_fun(statistic ~ style_sigfig) %>% 
  modify_caption("**Table 1. Patient Characteristic**")  %>%
  modify_header(label ~ "**Variable**") %>% 
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Survival Status**") %>%
  modify_footnote(all_stat_cols() ~ "Mean (SD) or Frequency (%)") %>%
  bold_labels() %>%
  as_gt()
```

+ the _t.test_ suggest that there is no significant difference in mean survival time and gtv between sex.

# Kaplan-Meier survival curves. 

```{r}
library(survminer)
fit <- survfit(Surv(time, status) ~ sex, data = data)
# Access to the sort summary table
summary(fit)$table

ggsurvplot(fit, data = data,
 surv.median.line = "hv", # Add medians survival

 # Change legends: title & labels
 legend.title = "Sex",
 legend.labs = c("Male", "Female"),
 # Add p-value and tervals
 pval = TRUE,

 conf.int = TRUE,
 # Add risk table
 risk.table = TRUE,
 tables.height = 0.2,
 tables.theme = theme_cleantable(),

 # Color palettes. Use custom color: c("#E7B800", "#2E9FDF"),
 # or brewer color (e.g.: "Dark2"), or ggsci color (e.g.: "jco")
 palette = "#Dark2",
 ggtheme = theme_bw(), # Change ggplot2 theme
 font.main = c(14, "bold", "darkblue"),
 font.x = c(12, "bold.italic", "darkblue"),
 font.y = c(12, "bold.italic", "darkblue"),
 font.tickslab = c(10, "plain", "darkgreen")
)
```

+ females tend to survive longer than men
+ the difference in survival time is not statistically significant at 5% level of significance

## difference in survival per diagnosis

```{r,echo=FALSE}
library(paletteer)
fit <- survfit(Surv(time, status) ~ diagnosis, data = data)

ggsurvplot(fit, data = data,
 surv.median.line = "hv", # Add medians survival

 # Change legends: title & labels
 legend.title = "diagnosis",
 # Add p-value and tervals
 pval = TRUE,

 conf.int = TRUE,
 # Add risk table
 risk.table = TRUE,
 tables.height = 0.2,
 tables.theme = theme_cleantable(),

 # Color palettes. Use custom color: c("#E7B800", "#2E9FDF"),
 # or brewer color (e.g.: "Dark2"), or ggsci color (e.g.: "jco")
 palette = "#Dark2",
 ggtheme = theme_bw(), # Change ggplot2 theme
 font.main = c(14, "bold", "darkblue"),
 font.x = c(12, "bold.italic", "darkblue"),
 font.y = c(12, "bold.italic", "darkblue"),
 font.tickslab = c(10, "plain", "darkgreen")
)
```

## difference in survival per Stereotactic

```{r,echo=FALSE}
fit <- survfit(Surv(time, status) ~ loc, data = data)

ggsurvplot(fit, data = data,
 surv.median.line = "hv", # Add medians survival

 # Change legends: title & labels
 legend.title = "location",
 # Add p-value and tervals
 pval = TRUE,

 conf.int = TRUE,
 # Add risk table
 risk.table = TRUE,
 tables.height = 0.2,
 tables.theme = theme_cleantable(),

 # Color palettes. Use custom color: c("#E7B800", "#2E9FDF"),
 # or brewer color (e.g.: "Dark2"), or ggsci color (e.g.: "jco")
 palette = "#Dark2",
 ggtheme = theme_bw(), # Change ggplot2 theme
 font.main = c(14, "bold", "darkblue"),
 font.x = c(12, "bold.italic", "darkblue"),
 font.y = c(12, "bold.italic", "darkblue"),
 font.tickslab = c(10, "plain", "darkgreen")
)
```

+ those with location at supratentorial have a longer survival time as compared to their counterparts
+ the difference in survival time is rather not statistically significant

# Now lets fit a model
## Logistic regression

```{r logistic_reg}
source(file = "R/helper.R")
data$status<-as.factor(data$status)

lr_mod <- parsnip::logistic_reg()|> 
   set_engine('glm') |> 
  set_mode("classification")  

model<-lr_mod |> 
  fit(status~. -time,data=data)

model %>%
  pluck("fit") %>%
  summary()
```
```{r,echo=FALSE}
model<-glm(status~. -time,data=data,family=binomial)
```

# perform a stepwise regression 
```{r}
stepmodel_logit <- stats::step(model, direction = "both", trace = FALSE)
summary(stepmodel_logit)
narative_model <- get_narative_model(model = stepmodel_logit, target = "status")
```
`r narative_model`

## **comments and interprettations**

+ from the logistic regression `diagnosis-HG glioma` and karnofsky index are statistically significant predictors of survival.
+ we speak of **odds** when we use logistic regression


## Now we fit a Cox proportional hazards model:

```{r}
cox_spec <- 
  proportional_hazards() %>% 
  set_engine("survival")

cox_fit <- cox_spec |> 
  fit(Surv(time, status) ~ ., data = BrainCancer)


cox_fit %>%
  pluck("fit") %>%
  summary()

```
`r narative_model`

## perform stepwise regression on `cox`

```{r , echo=FALSE}
cox_fit <- coxph(Surv(time, status) ~ ., data = BrainCancer)
```

```{r}
stepmodel_cox <- MASS::stepAIC(cox_fit, direction = "both", trace = FALSE)
summary(stepmodel_cox)
narative_model <- get_narative_model_cox(model = stepmodel_cox, target = "status")
```
`r narative_model`



