---
title: "Tidymodels in Healthcare"
author: "BonganI Ncube"
date: "2023-09-17"
slug: "tidymodels-in-healthcare"
categories:
- classification
- datascience
tags: []
subtitle: ''
summary: ''
authors: []
lastmod: "2023-09-17T22:30:07+02:00"
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
<script src="//yihui.org/js/math-code.js" defer></script>
<!-- Just one possible MathJax CDN below. You may use others. -->
<script defer
  src="//mathjax.rstudio.com/latest/MathJax.js?config=TeX-MML-AM_CHTML">
</script>



### Load packages

```{r setup, message=FALSE}

library(skimr)
library(tidyverse)
library(tidymodels)
library(scales)
library(ggridges)
library(patchwork)
library(gt)
library(xgboost)
library(ranger)
library(vip)
library(usemodels)
library(GGally)
library(glmnet)

```

### Load the Data

```{r, message=FALSE}
td_clean <- MLDataR::thyroid_disease
```

# Exploratory data analysis

As a strong visual learner, it helps me to see the data in both table and plot formats. Let's get started by looking at a few the summary statistics generated from the `skimr::skim()` combined with `summary()`. I will also use `glimpse()` to get a concise view of the data names, data types and the first few entries.

```{r}
td_clean %>% 
  skim() %>% 
  select(-(numeric.p0:numeric.p100)) %>%
  select(-(complete_rate)) %>% 
  summary()

td_clean %>% 
  glimpse()
```

```{r caption ='Loan data from lendingclub.com'}
td_clean %>% 
  head() %>% 
  gt() %>%
  tab_header(title = md("**First 6 rows of loan data from lendingclub.com**"))  %>%
  tab_options(container.height = 400,
              container.overflow.y = TRUE,
              heading.background.color = "#21908CFF",
              table.width = "75%",
              column_labels.background.color = "black",
              table.font.color = "black") %>%
  tab_style(style = list(cell_fill(color = "#35B779FF")),
            locations = cells_body())
```

The loan data consists of one character variable ('purpose' = loan type) and 13 numeric variables. Let's use `geom_col()` to plot the number of times each loan type is represented in the data.

```{r fig.cap = "Counts of each loan type", message=FALSE}
td_clean %>% 
  count(ThryroidClass) %>% 
  mutate(ThryroidClass = fct_reorder(ThryroidClass, n)) %>% 
  ggplot(aes(n, ThryroidClass, fill = ThryroidClass)) +
  geom_col() +
  labs(title = 'ThryroidClass by type',
       x = 'Count',
       y = '') +
  theme(legend.position = 'none') +
  tvthemes::scale_fill_avatar()

```


```{r clean_data}
td_clean <- td_clean[complete.cases(td_clean),]
dim(td_clean)
```

## Explanatory data analysis

```{r distribution_function}
# Get continuous variables only
subset <- td_clean %>% 
  dplyr::select(ThryroidClass, patient_age, TSH_reading, T3_reading,
                T4_reading, thyrox_util_rate_T4U_reading,
                FTI_reading)


# Bring in external file for visualisations
source('functions/visualisations.R')

# Use plot function
plot <- histoplotter(subset, ThryroidClass, 
                     chart_x_axis_lbl = 'Thyroid Class', 
                     chart_y_axis_lbl = 'Measures',boxplot_color = 'navy', 
                     boxplot_fill = '#89CFF0', box_fill_transparency = 0.2) 

# Add extras to plot
plot + ggthemes::theme_solarized() + theme(legend.position = 'none') + 
  scale_color_manual(values=c('negative' = 'red', 'positive' = 'blue'))


```

Numeric variables can be viewed as count distributions using `geom_histogram()`. Here, I use the `gather()` function to reshape the data into a long format and then `facet_wrap()` is used to display each histogram in a single figure.

```{r fig.cap = "Histograms of all numeric variables", message=FALSE}
td_clean %>% 
 dplyr::select(patient_age, TSH_reading, T3_reading,
                T4_reading, thyrox_util_rate_T4U_reading,
                FTI_reading) |>  
  gather() %>% 
  ggplot(aes(value, fill = key)) +
  geom_histogram(alpha = 0.7) +
  facet_wrap(~key, scales = 'free') +
  labs(title = 'Count distributions for numeric variables') +
  theme(legend.position = 'none')+
  tvthemes::scale_fill_avatar()
```

## Splitting the data

The data is split into training and test sets using the default (prop = 0.75) split. 5-fold cross validation is used to create resamples from the training set.

```{r divide_and_conquer}
td_clean <- td_clean %>% 
  dplyr::mutate(ThryroidClass = as.factor(ThryroidClass)) %>% 
  dplyr::select(-ref_src) %>% 
  drop_na()

# Split the dataset 
td_split <-initial_split(td_clean, 
                                   strata = ThryroidClass, 
                                   prop=0.9,
                                   breaks = 4)

train <- training(td_split)
test <- testing(td_split)

resamps <- vfold_cv(train, v = 5)
```

## Predicting loan type

To predict the loan type we will need to build a multiclass classification model. I am going to try three different models and compare the results to determine which model has the best predictive power.

### Model 1: XGboost

```{r}
#usemodels::use_xgboost(purpose ~ ., data = train)
train_rcp <- recipes::recipe(ThryroidClass ~ ., data=train) %>% 
  themis::step_smote(ThryroidClass, over_ratio = 0.97, neighbors = 3) %>%
  step_zv(all_predictors()) 

# Prep and bake the recipe so we can view this as a seperate data frame
training_df <- train_rcp %>% 
  prep() %>% 
  juice()

# Class imbalance resolved
class_imbalance_after_smote <- unclass(prop.table(table(training_df$ThryroidClass)))[1:2]

print(class_imbalance_after_smote)

xgboost_spec <-
  boost_tree(
    trees = 1000,
    tree_depth = tune(),
    min_n = tune(),
    mtry = tune(),
    sample_size = tune(),
    learn_rate = tune()
  ) %>%
  set_engine("xgboost") %>%
  set_mode("classification")

xgboost_workflow <-
  workflow() %>%
  add_recipe(train_rcp) %>%
  add_model(xgboost_spec)

# xgboost_grid_1 <- grid_regular(
#   parameters(xgboost_spec),
#   levels = 3 )

# xgboost_grid_2 <- grid_regular(
#   tree_depth(c(5L, 10L)),
#     min_n(c(10L, 40L)),
#     mtry(c(5L, 10L)),
#     learn_rate(c(-2, -1)),
#   sample_size=20
# )

set.seed(8390)
xgboost_tune <-
  tune_grid(xgboost_workflow, resamples = resamps, grid = 20)

xgboost_tune %>%
  autoplot()

xg_best_params <- xgboost_tune %>%
  select_best(metric = 'accuracy')

xg_best_spec <- xgboost_spec %>%
  finalize_model(xg_best_params)

xg_final_fit <- workflow() %>%
  add_recipe(train_rcp) %>%
  add_model(xg_best_spec) %>%
  last_fit(split = split)

xg_final_fit %>%
  collect_metrics()
```

### Model 2: Random forest

```{r}
# usemodels::use_ranger(purpose ~ ., data = train)

ranger_recipe <-
  recipe(formula = ThryroidClass ~ ., data=train)

ranger_spec <-
  rand_forest(mtry = tune(),
              min_n = tune(),
              trees = 1000) %>%
  set_mode("classification") %>%
  set_engine("ranger")

rand_forest(mode = "classification") %>%
  set_engine("ranger", importance = "impurity") %>%
  fit(ThryroidClass ~ ., data=train) %>%
  vip::vip(geom = 'point',
           aesthetics = list(color = 'midnightblue', size = 4)) +
  labs(title = 'test')

ranger_workflow <-
  workflow() %>%
  add_recipe(train_rcp) %>%
  add_model(ranger_spec)

# ranger_grid <- grid_regular(
#   parameters(ranger_spec),
#   mtry(range = c(10, 30)),
#   min_n(range = c(2, 8)),
#   levels = 3)

set.seed(57341)
ranger_tune <-
  tune_grid(ranger_workflow,
            resamples = resamps,
            grid = 20)

ranger_tune %>%
  autoplot()

rf_best_params <- ranger_tune %>%
  select_best(metric = 'accuracy')

rf_best_spec <- ranger_spec %>%
  finalize_model(rf_best_params)

rf_final_fit <- workflow() %>%
  add_recipe(train_rcp) %>%
  add_model(rf_best_spec) %>%
  last_fit(split = split)

rf_final_fit %>%
  collect_metrics()
```


```{r}
models <- c('xgboost', 'xgboost', 'random forrest', 'random forrest')
bind_rows(xg_final_fit %>% collect_metrics(), rf_final_fit %>% collect_metrics()) %>% 
  mutate(models = models) %>%
  select(models, everything(), -.config) %>% 
  gt() %>%
  tab_header(title = 'Model Results')  %>%
  tab_options(container.height = 400,
              container.overflow.y = TRUE,
              heading.background.color = "#21908CFF",
              table.width = "75%",
              column_labels.background.color = "black",
              table.font.color = "black") %>%
  tab_style(style = list(cell_fill(color = "#35B779FF")),
            locations = cells_body())
```

